---
title: "Chapter 5 The Many Variables and The Spurious Waffles"
author: "Trevor Williams"
date: "2026-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```

## 5.1 Spurious association

```{r 5.1}
# Load in data and standardize variables
library(rethinking)
data("WaffleDivorce")
d <- WaffleDivorce

# standardize variables
d$D <- standardize(d$Divorce)
d$M <- standardize(d$Marriage)
d$A <- standardize(d$MedianAgeMarriage)
```

Linear regression model for divorce rate ~ median age at marriage:

$$ D_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \beta_{A}A_{i} $$
$$ \alpha \sim \text{Normal}(0, 0.2) $$
$$ \beta_{A} \sim \text{Normal}(0,0.5) $$
$$ \sigma \sim \text{Exponential}(1) $$
Where $D_{i}$ is the standardized divorce rate for State $i$ and $A_{i}$ is State $i$'s standardized median age at marriage

```{r 5.2}
# See how large one stadnard deviation of age at marriage is
sd(d$MedianAgeMarriage)
```

```{r 5.3}
# Approximate posterior distribution
m5.1 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA * A,
    a ~ dnorm(0, 0.2),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d)
```

```{r 5.4}
# Simulate priors and plot lines for 2 standard deviations
set.seed(10)
prior <- extract.prior(m5.1)
mu <- link(m5.1, post=prior, data=list(A=c(-2,2)))
plot(NULL, xlim=c(-2,2), ylim=c(-2,2))
for (i in 1:50) lines( c(-2, 2), mu[i,], col=col.alpha("black", 0.4))
```

```{r 5.5}
# compute percentile interval of mean
A_seq <- seq(from = -3, to=3.2, length.out=30)
mu <- link(m5.1, data=list(A=A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# plot it all
plot(D ~ A, data=d, col=rangi2)
lines(A_seq, mu.mean, lwd=2)
shade(mu.PI, A_seq)
```

```{r 5.6}
# Similar regression for divorce rate ~ marriage rate
m5.2 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM * M,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d)

```


```{r}
precis(m5.1)
```

```{r}
precis(m5.2)
```
### 5.1.1

```{r 5.7}
# Draw the first dag in the section
library(dagitty)
dag5.1 <- dagitty("dag{ A -> D; A -> M; M -> D }")
coordinates(dag5.1) <- list( x=c(A=0,D=1,M=2), y=c(A=0,D=1,M=0))
drawdag(dag5.1)

```

### 5.1.2 Testable implications

```{r 5.8}
# Display implied conditional independencies from second DAG
DMA_dag2 <- dagitty('dag{ D <- A -> M}')
impliedConditionalIndependencies(DMA_dag2)
```

```{r}
# Check conditional independencies from first DAG
DMA_dag1 <- dagitty('dag{ D <- A -> M -> D}')
impliedConditionalIndependencies(DMA_dag1)
```

### 5.1.3 Multiple regression notation

Multiple regression model for divorce, marriage, and age:

$$ D_{i} \sim \text{Normal}(\mu_{i},\sigma) $$
$$ \mu_{i} = \alpha + \beta_{M}M_{i} + \beta_{A}A_{i} $$
$$ \alpha \sim \text{Normal}(0, 0.2) $$
$$ \beta_{M} \sim \text{Normal}(0, 0.5) $$
$$ \beta_{A} \sim \text{Normal}(0, 0.5) $$
$$ \sigma \sim \text{Exponential}(1) $$

### 5.1.4 Approximating the posterior

```{r 5.10}
# Fit multiple regression divorce rate model using quap
m5.3 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0, 0.2),
    bM <- dnorm(0, 0.5),
    bA <- dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)
precis(m5.3)
```

```{r 5.11}
# Plot posterior distributions for all three models, focusing on slope parameters
plot(coeftab(m5.1, m5.2, m5.3), par=c("bA", "bM"))
```

```{r 5.12}
# Simulating divorce example
N <- 50 # Number of simulated States
age <- rnorm(N) # sim A
mar <- rnorm(N, age) # sim A -> M
div <- rnorm(N, -age) # sim A -> D

d_sim <- list(A=age, M=mar, D=div)
sim_5.1 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA * A,
    a ~ dnorm(0, 0.2),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d_sim
)

sim_5.2 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM * M,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d_sim
)

sim_5.3 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0, 0.2),
    bM <- dnorm(0, 0.5),
    bA <- dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_sim
)

precis(sim_5.3)

# Plot posterior distributions for all three models, focusing on slope parameters
plot(coeftab(sim_5.1, sim_5.2, sim_5.3), par=c("bA", "bM"))
```

### 5.1.5 Plotting multivariate posteriors

#### 5.1.5.1 Predictor residual plots

To compute predictor residuals we need to suse the other predictor to model it, for marriage rate the model we need is:

$$ M_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \beta A_{i} $$
$$ \alpha \sim \text{Normal}(0,0.2) $$
$$ \beta \sim \text{Normal}(0.0.5) $$
$$ \sigma \sim \text{Exponential}(1) $$

```{r 5.13}
# Fit marriage rate as a function of median age at marriage
m5.4 <- quap(
  alist(
    M ~ dnorm(mu, sigma),
    mu <- a + bAM*A,
    a ~ dnorm(0, 0.2),
    bAM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)
```

```{r 5.14}
# Compute residual by subtracting the observed marriage rate from the predicted rate based on model
mu <- link(m5.4)
mu_mean <- apply(mu, 2, mean)
mu_resid <- d$M - mu_mean
```

#### 5.1.5.2 Posterior prediction plots

```{r 5.15}
# Posterior predictive check

# call link without specifying new data
# so it uses original data
mu <- link (m5.3)

# summarize samples across cases
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)

# simulate observations
# again no new data, so uses oriignal data
D_sim <- sim(m5.3, n=1e4)
D_PI <- apply(D_sim, 2, PI)
```

```{r 5.16}
# plot posterior predictive distributions for each sample

plot(mu_mean ~ d$D, col=rangi2, ylim=range(mu_PI),
     xlab="Observed divorce", ylab="Predicted divorce")
abline(a=0, b=1, lty=2)
for (i in 1:nrow(d)) lines(rep(d$D[i], 2), mu_PI[,i], col=rangi2)
```

```{r 5.17}
# Label points

plot(mu_mean ~ d$D, col=rangi2, ylim=range(mu_PI),
     xlab="Observed divorce", ylab="Predicted divorce")
abline(a=0, b=1, lty=2)
for (i in 1:nrow(d)) lines(rep(d$D[i], 2), mu_PI[,i], col=rangi2)

identify(x=d$D, y=mu_mean, labels=d$Loc)
```

```{r 5.18}
# Simulate scenario where a variable influences both the outcome and a spurious predictor
N <- 100 # Number of cases
x_real <- rnorm(N) # x_real as Gaussian with mean 0 and stddev 1
x_spur <- rnorm(N, x_real) # x_spur as Gaussian with mean=x_real
y <- rnorm(N, x_real) # y as Gaussian with mean=x_real
d <- data.frame(y, x_real, x_spur)
pairs(d)
```

```{r 5.19}
# Counterfactual plots, run quap with two regressions one altering A
data("WaffleDivorce")
d <- list()
d$A <- standardize(WaffleDivorce$MedianAgeMarriage)
d$D <- standardize(WaffleDivorce$Divorce)
d$M <- standardize(WaffleDivorce$Marriage)

m5.3_A <- quap(
  alist(
    ## A -> D <- M
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0,0.5),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1),
    ## A -> M
    M ~ dnorm(mu_M, sigma_M),
    mu_M <- aM + bAM*A,
    aM ~ dnorm(0, 0.2),
    bAM ~ dnorm(0,0.5),
    sigma_M ~ dexp(1)
  ),
  data = d
)
precis(m5.3_A)
```
```{r 5.20}
# define list of A values
A_seq <- seq(from=-2, to=2, length.out=30)
```

```{r 5.21}
# simulate observations from model to test intervention of A
# prep data
sim_dat <- data.frame(A=A_seq)

# Simulate M and then D, using A_seq
s <- sim(m5.3_A, data=sim_dat, vars=c("M", "D"))
```

```{r 5.22}
# PLot the predictions
plot(sim_dat$A, colMeans(s$D), ylim=c(-2,2), type="l",
     xlab="manipulated A", ylab="counterfactual D")
shade(apply(s$D, 2, PI), sim_dat$A)
mtext("Total counterfactual effect of A on D")
```

```{r}
# PLot the predictions
plot(sim_dat$A, colMeans(s$M), ylim=c(-2,2), type="l",
     xlab="manipulated A", ylab="counterfactual M")
shade(apply(s$M, 2, PI), sim_dat$A)
mtext("Total counterfactual effect A -> M")
```

```{r 5.23}
# See effected of increasing median age at marriage from 20 to 30
# new data frame, standardized to mean 26.1 and std dev 1.24
sim2_dat <- data.frame(A=(c(20,30)-26.1)/1.24)
s2 <- sim(m5.3_A, data=sim2_dat, vars=c("M", "D"))
mean(s2$D[,2]-s2$D[,1])
```

```{r 5.24}
# Couterfactual result of manipulating M
sim_dat <- data.frame(M=seq(from=-2,to=2, length.out=30), A=0)
s <- sim(m5.3_A, data=sim_dat, vars="D")

plot(sim_dat$M, colMeans(s), ylim=c(-2,2), type="l",
     xlab="manipulated M", ylab="counterfactual D")
shade(apply(s,2,PI), sim_dat$M)
mtext("Total counterfactual effect of M on D")
```

```{r 5.25}
# Simulating counterfactuals
A_seq <- seq(from=-2, to=2, length.out=30)
```

```{r 5.26}
# extract posterior samples and simulate M across values of A
post <- extract.samples(m5.3_A)
M_sim <- with(post, sapply(1:30 ,
                           function(i) rnorm(1e3, aM + bAM*A_seq[i], sigma_M)))
```

```{r 5.27}
# simulate D
D_sim <- with(post, sapply(1:30,
                           function(i) rnorm(1e3, a + bA*A_seq[i] + bM*M_sim[,i], sigma)))
```

## 5.2 Masked relationship

```{r 5.28}
# Load in milk data
data(milk)
d <- milk
str(d)
```

```{r 5.29}
# Standardize variables
d$K <- standardize(d$kcal.per.g)
d$N <- standardize(d$neocortex.perc)
d$M <- standardize(log(d$mass))
```

First model kilocalories ~ neocortex percent

$$ K_{i} \sim \text{Normal}(\mu_{i], \sigma}) $$
$$ \mu_{i} = \alpha + \beta_{N}N_{i} $$

```{r 5.30}
# Try to model bivariate model above
m5.5_draft <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 1),
    bN ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = d
)
```

```{r 5.32}
# Drop missing values - Complete Case Analysis
dcc <- d[complete.cases(d$K,d$N,d$M),]
```


```{r 5.33}
# Create model
m5.5_draft <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 1),
    bN ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dcc
)
```

```{r 5.34}
# Evaluate priors using prior predictive simulations
prior <- extract.prior(m5.5_draft)
xseq <- c(-2, 2)
mu <- link(m5.5_draft, post=prior, data=list(N=xseq))
plot(NULL, xlim=xseq, ylim=xseq)
for (i in 1:50) lines(xseq, mu[i,], col=col.alpha("black", 0.3))
```

```{r 5.35}
# Tighten priors

m5.5 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = dcc
)

# Evaluate new priors using prior predictive simulations
prior <- extract.prior(m5.5)
xseq <- c(-2, 2)
mu <- link(m5.5, post=prior, data=list(N=xseq))
plot(NULL, xlim=xseq, ylim=xseq)
for (i in 1:50) lines(xseq, mu[i,], col=col.alpha("black", 0.3))
```

```{r 5.36}
# Evaluate posterior
precis(m5.5)
```

```{r 5.37}
# Plot posterior
xseq <- seq(from=min(dcc$N)-0.15, to=max(dcc$N)+0.15, length.out=30)
mu <- link(m5.5, data=list(N=xseq))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
plot(K ~ N, data=dcc)
lines(xseq, mu_mean, lwd=2)
shade(mu_PI, xseq)
```

```{r 5.38}
# Now construct model of log(female body mass) to kilocalories
m5.6 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bM*M,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = dcc
)
precis(m5.6)
```

```{r}
# plot posterior of log body mass model
m_seq <- seq(min(dcc$M) - 0.15, max(dcc$M) + 0.15, length.out=30)
mu <- link(m5.6, data=list(M=m_seq))
mu_mean = apply(mu, 2, mean)
mu_PI = apply(mu, 2, PI)
plot(K ~ M, data=dcc)
lines(m_seq, mu_mean, col=rangi2)
shade(mu_PI, m_seq)
```

Multivariate model with both predictors

$$ K_{i} \sim \text{Normal}(\mu_{i], \sigma}) $$
$$ \mu_{i} = \alpha + \beta_{N}N_{i} + \beta_{M}M_{i} $$
$$ \alpha \sim \text{Normal}(0, 0.2) $$
$$ \beta_{N} \sim \text{Normal}(0,0.5) $$
$$ \beta_{M} \sim \text{Normal}(0,0.5) $$
$$ \sigma \sim \text{Exponential}(1) $$
```{r 5.39}
# Approximate  posterior of multivariate model
m5.7 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N + bM*M,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ),
  data=dcc
)

precis(m5.7)
```

```{r 5. 40}
# Plot difference in posteriors between univariate and multivariate models
plot(coeftab(m5.5, m5.6, m5.7), pars=c("bM", "bN"))
```

```{r}
pairs(~K + M + N, dcc)
```

```{r 3.41}
# Counterfactual plots

xseq <- seq(from=min(dcc$M)-0.15, to=max(dcc$M)+0.15, length.out=30)
mu <- link(m5.7, data=data.frame(M=xseq, N=0))
mu_mean <- apply(mu,2,mean)
mu_PI <- apply(mu,2,PI)
plot(NULL, xlim=range(dcc$M), ylim=range(dcc$K))
lines(xseq, mu_mean, lwd=2)
shade(mu_PI, xseq)
```

```{r}
xseq <- seq(from=min(dcc$N)-0.15, to=max(dcc$N)+0.15, length.out=30)
mu <- link(m5.7, data=data.frame(N=xseq, M=0))
mu_mean <- apply(mu,2,mean)
mu_PI <- apply(mu,2,PI)
plot(NULL, xlim=range(dcc$N), ylim=range(dcc$K))
lines(xseq, mu_mean, lwd=2)
shade(mu_PI, xseq)
mtext("")
```

```{r 3.42}
# simulating a masking relationship first dag
# M -> K <- N
# M -> N
n <- 100
M <- rnorm(n)
N <- rnorm(n, M)
K <- rnorm(n, N-M)
d_sim <- data.frame(K=K, N=N, M=M)

m5.5_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_sim
)

m5.6_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bM*M,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_sim
)

m5.7_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N + bM*M,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ),
  data=d_sim
)

plot(coeftab(m5.5_dagsim, m5.6_dagsim, m5.7_dagsim), pars=c("bM", "bN"))

```

```{r 3.43}
# Simulating the other dags

# M -> K <- N
# N -> M
n <- 100
N <- rnorm(n)
M <- rnorm(n, N)
K <- rnorm(n, N-M)
d_sim2 <- data.frame(K=K,N=N,M=M)

# M -> K <- N
# M <- U -> N
n <- 100
U <- rnorm(n)
N <- rnorm(n, U)
M <- rnorm(n, U)
K <- rnorm(n, N-M)
d_sim3 <- data.frame(K=K,N=N,M=M)

m5.5_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_sim3
)

m5.6_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bM*M,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_sim3
)

m5.7_dagsim <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N + bM*M,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ),
  data=d_sim3
)

plot(coeftab(m5.5_dagsim, m5.6_dagsim, m5.7_dagsim), pars=c("bM", "bN"))
```

```{r 5.44}
# Get Markov Equivalence set
dag5.7 <- dagitty("dag{
                  M -> K <- N
                  M -> N}")
coordinates(dag5.7) <- list(x=c(M=0,K=1,N=2), y=c(M=0.5,K=1,N=0.5))
MElist <- equivalentDAGs(dag5.7)
drawdag(MElist)
```

## 5.3 Categorical variables

### 5.3.1 Binary categories

```{r 5.45}
# Look at Howell data variables
data(Howell1)
d <- Howell1
str(d)
```

Height model with only sex variable as an indicator variable:

$$ h_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \beta_{m}m_{i} $$
$$ \alpha \sim \text{Normal}(178, 20) $$
$$ \beta_{m} \sim \text{Normal}(0, 10) $$

$$ \sigma \sim \text{Uniform}(0,50) $$

```{r 5.46}
# Evaluate prior distributions
mu_female <- rnorm(1e4,178,20)
mu_male <- rnorm(1e4,178,20) + rnorm(1e4,0,10)
precis(data.frame(mu_female, mu_male))
```

```{r 5.47}
# Construct index variable
d$sex <- ifelse(d$male == 1, 2, 1)
str(d$sex)
```

height model with sex as an index variable

$$ h_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha_{SEX[i]} $$
$$ \alpha_{j} \sim \text{Normal}(178, 20) \text{for} j=1..2 $$
$$ \sigma \sim \text{Uniform}(0,50) $$

```{r 5.48}
# approximate posterior for sex index variable model
m5.8 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a[sex],
    a[sex] ~ dnorm(178, 20),
    sigma ~ dunif(0,50)
  ),
  data=d
)
precis(m5.8, depth=2)
```

```{r 5.49}
# Extract samples from posterior and compute difference in height between sexes (a contrast)
post <- extract.samples(m5.8)
post$diff_fm <- post$a[,1] - post$a[,2]
precis(post, depth=2)
```

### 5.3.2 Many categories

```{r 5.50}
# Return to milk data
data(milk)
d <- milk
levels(d$clade)
```

```{r 5.51}
# coerce factor to integer
d$clade_id <- as.integer(d$clade)
```

Model to measure average milke energy in each clade:

$$ K_{i} \sim \text{Normal}(\mu_{i},\sigma) $$
$$ \mu_{i} = \alpha_{CLADE[i]} $$
$$ \alpha_{j} \sim \text{Normal}(0,0.5) \text{ for } j=1..4 $$
$$ \sigma \sim \text{Exponential}(1) $$
```{r 5.52}
# approximate posterior for clade model
d$K <- standardize(d$kcal.per.g)
m5.9 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a[clade_id],
    a[clade_id] ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d
)
labels <- paste("a[", 1:4, "]:", levels(d$clade), sep="")
plot(precis(m5.9, depth=2, pars="a"), labels=labels, xlab='expected kcal (std)')
```

```{r 5.53}
# Ad another arbitrary grouping variable
set.seed(63)
d$house <- sample(rep(1:4, each=8), size=nrow(d))
```

```{r 5.54}
# include new grouping variable into model
m5.10 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a[clade_id] + h[house],
    a[clade_id] ~ dnorm(0, 0.5),
    h[house] ~ dnorm(0,0.5),
    sigma ~ dexp(1)
  ),
  data=d
)

precis(m5.10, depth=2)
```






