---
title: "Chapter 11: God Spiked the Integers"
author: "Trevor Williams"
date: "2026-08-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 11.1 Binomial regression

The binomial distribution is denoted:

$$ y \sim \text{Binomial}(n,p) $$
Where $y$ is a count (zero or a positive whole number), $p$ is the probability any particular "trial" is a success, and $n$ is the number of trials.

Two flavors of GLM that use binomial probability functions:

 1. Logistic Regression: data are organized into single-trial cases, so outcome can only take a 0 or a 1
 2. Aggregated Binomial Regression: outcome can take teh value zero or any positive integer up to $n$
 

### 11.1.1 Logistic regression: Prosocial Chimpanzees

Chimpanzee prosocial example: goal is to estimate the interaction between condition (presence of absence of another animal) and option (which side is prosocial)


```{r 11.1}
library(rethinking)
data(chimpanzees)
d <- chimpanzees
```
We are going to model `pulled_left` as the outcome to predict, with `prosoc_left` and `condition` as predictor variables

Four combinations of `prosoc_left` and `condition` we need to infer:

  1. `prosoc_left` = 0 and `condition` = 0: two food items on the right and no partner
  2. `prosoc_left` = 1 and `condition` = 0: two food items on the left and no partner
  3. `prosoc_left` = 0 and `condition` = 1: two food items on the right and partner present
  4. `prosoc_left` = 1 and `condition` = 1: two food items on the left and partner present
  
We will create an index fariable to index the combinations above:

```{r 11.2}
d$treatment <- 1 + d$prosoc_left + 2*d$condition
```

verify by using crosstab


```{r 11.3}
xtabs( ~ treatment + prosoc_left + condition, d)
```

Before building the model, let's consider a runt of logistic regression with just a single $\alpha$ paramter:

$$ L_1 \sim \text{Binomial}(1, p_i) $$
$$ \text{logit}(p_i) = \alpha $$
$$ \alpha \sim \text{Normal}(0, \omega) $$

Need to pick a value for $\omega$ lets start with $\omega = 10$

```{r 11.4}

m11.1 <- quap(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a,
    a ~ dnorm(0, 10)
  ), data=d
)

```

Now sample from the prior

```{r 11.5}

set.seed(1999)
prior <- extract.prior(m11.1, n=1e4)

```


Next we need to convert the parameter to the outcome scale using the inverse-link function

```{r 11.6}

p <- inv_logit(prior$a)
dens(p, adj=0.1)

```

Now do the same thing for treatment effects

```{r 11.7}

m11.2 <- quap(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a + b[treatment],
    a ~ dnorm(0, 1.5),
    b[treatment] ~ dnorm(0, 10)
  ), data=d
)
set.seed(1999)
prior <- extract.prior(m11.2, n=1e4)
p <- sapply(1:4, function(k) inv_logit(prior$a + prior$b[,k]))

```

and plot it 

```{r 11.8}

dens(abs(p[,1] - p[,2]), adj=0.1)

```

look at average prior distance using more informative prior

```{r 11.9}

m11.3 <- quap(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a + b[treatment],
    a ~ dnorm(0, 1.5),
    b[treatment] ~ dnorm(0, 0.5)
  ), data=d
)
set.seed(1999)
prior <- extract.prior(m11.3, n=1e4)
p <- sapply(1:4, \(k) inv_logit(prior$a + prior$b[,k]))
mean( abs(p[,1] - p[,2]))

```

Prepare data lists to run HMC

```{r 11.10}

# trimmed data list
dat_list <- list(
  pulled_left = d$pulled_left,
  actor = d$actor,
  treatment = as.integer(d$treatment)
)

```


Fit the model

```{r 11.11}

m11.4 <- ulam(
  alist(
    pulled_left ~ dbinom(1, p), 
    logit(p) <- a[actor] + b[treatment],
    a[actor] ~ dnorm(0, 1.5),
    b[treatment] ~ dnorm(0, 0.5)
  ), data=dat_list, chains=4, log_lik=TRUE, cores=4
)
precis(m11.4, depth=2)

```

Look at parameters on the outcome scale

```{r 11.12}
post <- extract.samples(m11.4)
p_left <- inv_logit(post$a)
plot(precis(as.data.frame(p_left)), xlim=c(0, 1))
```

Now look at treatment effects

```{r 11.13}

labs <- c("R/N", "L/N", "R/P", "L/P")
plot(precis(m11.4, depth=2, pars="b"), labels=labs)

```

calculate differences between no-partner/partner 

```{r 11.14}

diffs <- list(
  db13 = post$b[,1] - post$b[,3],
  db24 = post$b[,2] - post$b[,4]
)
plot(precis(diffs))

```

Consider posterior prediction check. First summarize the porportions of left pulls for each actor in each treatment

```{r 11.15}

pl <- by(d$pulled_left, list(d$actor, d$treatment), mean)
pl[1,]

```

Plot raw data

```{r 11.16}

plot(NULL, xlim=c(1,28), ylim=c(0,1), xlab="",
     ylab = "Proportion left lever", xaxt="n", yaxt="n")
axis(2, at=c(0, 0.5, 1), labels = c(0, 0.5, 1))
abline(h=0.5, lty=2)
for (j in 1:7) abline(v=(j-1)*4+4.5, lwd=0.5)
for (j in 1:7) text( (j-1)*4+2.5, 1.1, concat("actor ", j), xpd=TRUE)
for (j in (1:7)[-2]) {
  lines( (j-1)*4+c(1,3), pl[j,c(1,3)], lwd=2, col=rangi2)
  lines( (j-1)*4+c(2,4), pl[j,c(2,4)], lwd=2, col=rangi2)
}
points(1:28, t(pl), pch=16, col="white", cex=1.7)
points(1:28, t(pl), pch=c(1,1,16,16), col=rangi2, lwd=2)
yoff <- 0.01
text(1, pl[1,1]-yoff, "R/N", pos=1, cex=0.8)
text(2, pl[1,2]-yoff, "L/N", pos=3, cex=0.8)
text(3, pl[1,3]-yoff, "R/P", pos=1, cex=0.8)
text(4, pl[1,4]-yoff, "L/P", pos=3, cex=0.8)
mtext("observed proportions\n")

```

Get posterior predictions

```{r 11.17}
dat <- list(actor=rep(1:7, each=4), treatment=rep(1:4,times=7))
p_post <- link(m11.4, data=dat)
p_mu <- apply(p_post, 2, mean)
p_ci <- apply(p_post, 2, PI)
```

Get index variables for interaction model

```{r 11.18}
d$side <- d$prosoc_left + 1 # right 1, left 2
d$cond <- d$condition + 1 # no partner 1, partner 2
```


```{r 11.19}

dat_list2 <- list(
  pulled_left = d$pulled_left,
  actor = d$actor,
  side = d$side,
  cond = d$cond
)
m11.5 <- ulam(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a[actor] + bs[side] + bc[cond],
    a[actor] ~ dnorm(0, 1.5),
    bs[side] ~ dnorm(0, 0.5),
    bc[cond] ~ dnorm(0, 0.5)
  ), data = dat_list2, chains = 4, cores = 4, log_lik = TRUE
)

```

compare two modles with PSIS

```{r 11.20}

compare(m11.5, m11.4, func=PSIS)

```


### 11.1.2 Relative shark and absolute deer

Discussed relative vs. absolute effects. Relative effects are exponentiating the betas

### 11.1.3 Aggregated binomial: Chimpanzees again, condensed

Data can be condensed into counts of left pulls

to calculate the number of times each chimpanzee pulled the left-hand lever, for each combination of predictor values:

```{r 11.24}

data(chimpanzees)
d <- chimpanzees
d$treatment <- 1 + d$prosoc_left + 2*d$condition
d$side <- d$prosoc_left + 1 # right 1, left 2
d$cond <- d$condition + 1 # no partner 1, partner 2
d_aggregated <- aggregate(
  d$pulled_left,
  list(treatment=d$treatment, actor=d$actor,
       side=d$side, cond=d$cond),
  sum
)

colnames(d_aggregated)[5] <- "left_pulls"

```

Now define the model

```{r}

```

