---
title: "Chapter 4 Geocentric Models"
output: html_notebook
---

```{r}
library(rethinking)
```

## 4.1 Why normal distributions are normal

### 4.1.1 Normal by addition

```{r 4.1}
# Simulate the coin flip and step example
# Any process htat adds together random values form teh same distribution
# converges to a normal

pos <- replicate(1000, sum(runif(16,-1,1)))
plot(density(pos))
```

### 4.1.2 Normal by multiplication
```{r 4.2}
# Example: Growth rate by 12 genes that is multiplicative
prod(1 + runif(12,0,0.1))
```

```{r 4.3}
# Now get 10000 random products
growth <- replicate(10000, prod(1 + runif(12, 0, 0.1)))
dens(growth, norm.comp=TRUE)
```

```{r 4.4}
# Compare big vs. small effects - small converges to normal better
big <- replicate(10000, prod(1 + runif(12,0,0.5)))
small <- replicate(10000, prod(1 + runif(12, 0, 0.01)))
dens(big, norm.comp = TRUE)
```

```{r}
dens(small, norm.comp=TRUE)
```

### 4.1.3 Normal by log-multiplication

```{r 4.5}
# Adding logs is equivalent to multiplying the original numbers
log.big <- replicate(10000, log(prod(1 + runif(12,0,0.5))))
dens(log.big, norm.comp = TRUE)
```

### 4.1.4 Using Gaussian distributions

No code snippets

## 4.2 A language for describing models

### 4.2.1 Re-describing the globe tossing model

Model description of globe-tossing model

$$ W \sim \text{Binomial}(N, p)$$
$$ p \sim \text{Uniform}(0,1) $$

Where $W$ is the observed count of water, $N$ is the total number of tosses, and $p$ is the proportion of water on the globe.

```{r 4.6}
# Use R to approximate equation specified in overthinking box
w <- 6; n <- 9;
p_grid <- seq(from=0, to=1, length.out=100)
posterior <- dbinom(w,n,p_grid)*dunif(p_grid,0,1)
posterior <- posterior/sum(posterior)

```

## 4.3 Gaussian model of height

### 4.3.1 The data

```{r 4.7}
# Load in data
library(rethinking)
data("Howell1")
d <- Howell1
```

```{r 4.8}
# Examine data features
str(d)
```

```{r 4.9}
# Examine data using rethinking precis summary function
precis(d)
```


```{r 4.10}
# Examine height data from dataframe
d$height
```

```{r 4.11}
# Filter to adults only
d2 <- d[ d$age >= 18, ]
```

### 4.3.2 The model

```{r}
dens(d2$height)
```

Likelihood:
$$ h_{i} \sim \text{Normal}(\mu, \sigma)$$
Priors:
$$ \mu \sim \text{Normal}(178, 20) $$
$$ \sigma \sim \text{Uniform}(0, 50) $$

```{r 4.12}
# Plot mu prior
curve(dnorm(x, 178, 20), from=100, to=250)
```

```{r 4.13}
# Plot sigma prior
curve(dunif(x, 0, 50), from=-10, to=60)
```

```{r 4.14}
# Conduct prior predictive simulations to see what priors imply
sample_mu <- rnorm(1e4, 178, 20) # Simulate 1000 'observations' from mu prior
sample_sigma <- runif(1e4, 0, 50) # Simulate 1000 'observations' from sigma prior
prior_h <- rnorm(1e4, sample_mu, sample_sigma) # Simulate height using simulated values from priors
dens(prior_h)
```


```{r 4.15}
# Use a larger sigma to compare prior predictive distributions
sample_mu <- rnorm(1e4, 178, 100)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
# Notice that it assigns plausibility to heights less than 0 and greater than 400
```

### 4.3.3 Grid approximation of teh posterior distribution

```{r 4.16}
# Approximate posterior distribution for height given data using grid approximation
mu.list <- seq(from=150, to=160, length.out=200)
sigma.list <- seq(from=7, to=9, length.out=200)
post <- expand.grid(mu=mu.list, sigma=sigma.list) # Get all combinations of mu and sigma
post$LL <- sapply(1:nrow(post), function(i) sum(dnorm(d2$height, post$mu[i], post$sigma[i], log=TRUE))) # calculate likelihood across all values
post$prod <- post$LL + dnorm(post$mu, 178, 20, TRUE) + dunif(post$sigma, 0, 50, TRUE) # multiply likelihood and prior
post$prob <- exp(post$prod - max(post$prod)) # Calculate relative posterior probabilities (to avoid rounding to zero)
```

```{r 4.17}
# Create contour plot of posterior
contour_xyz(post$mu, post$sigma, post$prob)
```


```{r 4.18}
# plot simple heat map of posterior
image_xyz(post$mu, post$sigma, post$prob)
```

### 4.3.4 Sampling from the posterior

```{r 4.19}
# sample parameter values in proportion to their posterior probabilities
sample.rows <- sample(1:nrow(post), size=1e4, replace=TRUE, prob=post$prob)
sample.mu <- post$mu[sample.rows]
sample.sigma <- post$sigma[sample.rows]
```

```{r 4.20}
# Plot posterior samples
plot(sample.mu, sample.sigma, cex=0.5, pch=16, col=col.alpha(rangi2, 0.1))
```

```{r 4.21}
# Get marginal posterior densities of mu and sigma
dens(sample.mu)
dens(sample.sigma)
```

```{r 4.22}
# Summarize widths of densities using posterior compatibility intervals
PI(sample.mu)
PI(sample.sigma)

```

```{r 4.23}
# Sample heights to see how a smaller number of data points affects posterior
d3 <- sample(d2$height, size=20)
```

```{r}
# Recalculate grid approximated posterior using d3 instead of d2
mu.list <- seq(from=150, to=170, length.out=200)
sigma.list <- seq(from=4, to=20, length.out=200)
post2 <- expand.grid(mu=mu.list, sigma=sigma.list)
post2$LL <- sapply(1:nrow(post2), function(i) sum(dnorm(d3, mean=post2$mu[i], sd=post2$sigma[i], log=TRUE)))
post2$prod <- post2$LL + dnorm(post2$mu, 178, 20, TRUE) + dunif(post2$sigma, 0, 50, TRUE)
post2$prob <- exp(post2$prod - max(post2$prod))
sample2.rows <- sample(1:nrow(post2), size=1e4, replace=TRUE, prob=post2$prob)
sample2.mu <- post2$mu[sample2.rows]
sample2.sigma <- post2$sigma[sample2.rows]
plot(sample2.mu, sample2.sigma, cex=0.5, col=col.alpha(rangi2, 0.1), xlab='mu', ylab='sigma', pch=16)
```

```{r}
# Examine marginal posterior density for sigma, averaging over mu
dens(sample2.sigma, norm.comp=TRUE)
```

### 4.3.5 Finding the posterior distribution with quap

```{r 4.26}
# Re-initialize data
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18, ]
```

```{r 4.27}
# Put model description into an `alist`
flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(178, 20),
  sigma ~ dunif(0, 50)
)
```

```{r 4.28}
# Fit model to data using `quap`

m4.1 <- quap(flist, data=d2)
```

```{r 4.29}
# Evaluate the posterior using `precis`
precis(m4.1)
```

```{r 4.30}
# Give a specified starting point for `quap`
start <- list(
  mu=mean(d2$height),
  sigma=sd(d2$height)
)
m4.1 <- quap(flist, data=d2, start=start)
precis(m4.1)
```


```{r 4.31}
# Use a more "informative" prior and see how it affects posterior
m4.2 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ dnorm(178, 0.1), # Narrow sd for mean height
    sigma ~ dunif(0, 50)
  ), data=d2
)
precis(m4.2)

# Notice that the posterior estimate for mu is close to the prior value
# Causing the sigma posterior to become much wider
```

### 4.3.6 Sampling from a quap

```{r 4.32}
# Examine the variance-covariance matrix for the posterior
vcov(m4.1)
```

```{r 4.33}
# Decomposition of variance-covariance matrix into vectors of individual variances
# and correlation between parameters
diag(vcov(m4.1))
cov2cor(vcov(m4.1))
```

```{r 4.34}
# Sample the posterior (sampling a multivariate Gaussian distribution)
post <- extract.samples(m4.1, n=1e4)
head(post)
```


```{r 4.35}
precis(post)
```

```{r 4.36}
# extract.samples under the hood
library(MASS)
post <- mvrnorm(n=1e4, mu=coef(m4.1), Sigma=vcov(m4.1))
```

## 4.4 Linear Prediction

```{r 4.37}
# Before creating a predictive model, plot height and weight against each other
data("Howell1"); d <- Howell1; d2 <- d[d$age >= 18,]
plot(d2$height ~ d2$weight)
```

### 4.4.1 The linear model strategy

$$ h_{i} \sim \text{Normal}(\mu_{i},\sigma) $$
$$ \mu_{i} = \alpha + \beta(x_{i}-\bar{x}) $$
$$ \alpha \sim \text{Normal}(178, 20) $$
$$ \beta \sim \text{Normal}(0, 10) $$
$$ \sigma \sim \text{Uniform}(0,50) $$

```{r 4.38}
# Simulate prior predictive distribution
set.seed(2971)
N <- 100 # Get 100 alpha and beta parameters
a <- rnorm(N, 178, 20)
b <- rnorm(N, 0, 10)
```

```{r 4.39}
# Plot lines
plot(NULL, xlim=range(d2$weight), ylim=c(-100,400),
     xlab='weight', ylab='height')
abline(h=0, lty=2)
abline(h=272, lty=1, lwd=0.5)
mtext("b ~ dnorm(0,10")
xbar <- mean(d2$weight)
for (i in 1:N) curve(a[i] + b[i]*(x-xbar),
                     from=min(d2$weight), to=max(d2$weight), add=TRUE,
                     col=col.alpha("black", 0.2))
```

```{r 4.41}
# Define beta as log-normal(0,1) instead to ensure the relationship is positive
b <- rlnorm(1e4, 0, 1)
dens(b, xlim=c(0,5), adj=0.1)

```

```{r}
# Re-do prior predictive simulation using log normal beta prior
set.seed(2971)
N <- 100
a <- rnorm(N, 178, 20)
b <- rlnorm(N, 0, 1)

# Plot lines
plot(NULL, xlim=range(d2$weight), ylim=c(-100,400),
     xlab='weight', ylab='height')
abline(h=0, lty=2)
abline(h=272, lty=1, lwd=0.5)
mtext("b ~ dnorm(0,10")
xbar <- mean(d2$weight)
for (i in 1:N) curve(a[i] + b[i]*(x-xbar),
                     from=min(d2$weight), to=max(d2$weight), add=TRUE,
                     col=col.alpha("black", 0.2))
```

### 4.4.2 Finding the posterior distribution

Better model with log-normal Beta prior

$$ h_{i} \sim \text{Normal}(\mu_{i},\sigma) $$
$$ \mu_{i} = \alpha + \beta(x_{i}-\bar{x}) $$
$$ \alpha \sim \text{Normal}(178, 20) $$
$$ \beta \sim \text{Log-Normal}(0, 1) $$
$$ \sigma \sim \text{Uniform}(0,50) $$
```{r 4.42}
# Fit linear predictive model

# load in data agin, since it's a long way back
library(rethinking)
data("Howell1"); d <- Howell1; d2 <- d[d$age >= 18, ]

# define the average weight, x-bar
xbar <- mean(d2$weight)

# fit model 
m4.3 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data=d2
)
```

```{r 4.43}
# Another way to code the same model without the log-normal distribution
m4.3b <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + exp(log_b)*(weight - xbar),
    a ~ dnorm(178, 20),
    log_b ~ dnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data = d2
)
```

### 4.4.3 Interpreting the posterior distribution

#### 4.4.3.1 Tables of marginal distributions

```{r 4.44}
# Get marginal posterior distribution information in table form
precis(m4.3)
```

```{r 4.45}
# Get variance-covariance table of parameters
round(vcov(m4.3), 3)
```

```{r}
pairs(m4.3)
```

#### 4.4.3.2 Plotting posterior inference against the data

```{r 4.46}
# Plot the line using the mean values of the parameter and compare against raw data
plot (height ~ weight, data=d2, col=rangi2)
post <- extract.samples(m4.3)
a_map <- mean(post$a)
b_map <- mean(post$b)
curve(a_map + b_map*(x-xbar), add=TRUE)
```

#### 4.4.3.3 Adding uncertainty around the mean

```{r 4.47}
# Posterior distribution contains lines at all values of alpha and beta
post <- extract.samples(m4.3)
post[1:5,]
```

```{r 4.48}
# Refit the model to only 10 individuals to see how that affects variance in lines
N <- 10
dN <- d2[1:N, ]
mN <- quap(
    alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data = dN
)
```


```{r 4.49}
# Plot 20 lines from the posterior to look at uncertainty
# extract 20 samples form the posterior
post <- extract.samples(mN, n=20)

# display raw data and sample size
plot(dN$weight, dN$height,
     xlim=range(d2$weight), ylim=range(d2$height),
     col=rangi2, xlab="weight", ylab="height")
mtext(concat("N = ", N))

# Plot the lines, with transparency
for (i in 1:20) {
  curve(post$a[i] + post$b[i]*(x-mean(dN$weight)),
        col=col.alpha("black", 0.3), add=TRUE)
}
```

```{r}
par(mfrow=c(2,2))

for(n in c(10, 50, 150, 352)) {
  dN <- d2[1:n, ]
  mN <- quap(
    alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data = dN
  )
  
  # Plot 20 lines from the posterior to look at uncertainty
  # extract 20 samples form the posterior
  post <- extract.samples(mN, n=20)
  
  # display raw data and sample size
  plot(dN$weight, dN$height,
       xlim=range(d2$weight), ylim=range(d2$height),
       col=rangi2, xlab="weight", ylab="height")
  mtext(concat("N = ", N))
  
  # Plot the lines, with transparency
  for (i in 1:20) {
    curve(post$a[i] + post$b[i]*(x-mean(dN$weight)),
          col=col.alpha("black", 0.3), add=TRUE)
  }
}
```

#### 4.4.3.4 Plotting regression intevals and contours

```{r 4.50} 
# Get 10000 values of height for an individual who weighs 50 kilograms
post <- extract.samples(m4.3)
mu_at_50 <- post$a + post$b * (50 - xbar)
```

```{r 4.51}
# Plot density of samples at 50 
dens(mu_at_50, col=rangi2, lwd=2, xlab="mu|weight=50")
```

```{r 4.52}
# Get 89% compatibility interval
PI(mu_at_50, prob=0.89)
```

```{r 4.53}
# Get posterior distribution of mu for every case (individual) in the dataset
mu <- link(m4.3)
str(mu)
```
```{r 4.54}
# Get posterior distributions of mu for every unique weight value

# define sequence of weights to compute predictions for
# these values will be on the horizontal axis
weight.seq <- seq(from=25, to=70, by=1)

# use link to compute mu
# for each sample from posterior
# and for each weight in weight.seq
mu <- link(m4.3, data=data.frame(weight=weight.seq))
str(mu)
```
 
```{r 4.55}
# Plot distribution of mu values at each weight

plot(height ~ weight, d2, type='n')

# loop over samples and plot each mu value
for(i in 1:100) {points(weight.seq, mu[i,], pch=16, col=col.alpha(rangi2,0.1))}

```

```{r 4.56}
# summarize the distribution of mu
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob=0.89)
data.frame(weight = weight.seq, mu.mean, t(mu.PI))
```

```{r}
# plot raw data
# fading out points to make line and interval more visible
plot(height~weight, data=d2, col=col.alpha(rangi2,0.5))

# Plot the MAP line, aka the mean mu for each weight
lines(weight.seq, mu.mean)

# plot a shaded region for 89% PI
shade(mu.PI, weight.seq)
```

```{r 4.58}
# `link` function under the hood
post <- extract.samples(m4.3)
mu.link <- function(weight) post$a + post$b * (weight-xbar)
weight.seq <- seq(from=25, to=70, by=1)
mu <- sapply(weight.seq, mu.link)
mu.mean <- apply(mu, 2, mean)
mu.CI <- apply(mu, 2, PI, prob=0.89)
```

#### 4.4.3.5 Prediction intervals

```{r 4.59}
# Simulate actual heights instead of mean heights embodying the uncertainty in the 
# posterior as well as teh uncertainty in the Gaussian distribution of heights
# using `sim`

sim.height <- sim(m4.3, data=list(weight=weight.seq))
str(sim.height)
```

```{r 4.60}
# summarize simulated heights
height.PI <- apply(sim.height, 2, PI, prob=0.89)
```

```{r}
# plot everything together
# plot raw data
plot(height ~ weight, data=d2, col=col.alpha(rangi2, 0.5))

# plot the MAP mu estimates
lines(weight.seq, mu.mean)

# draw HPDI region for line
mu.HPDI <- apply(mu, 2, HPDI)
shade(mu.HPDI, weight.seq)

# draw PI region for simulated weights
shade(height.PI, weight.seq)
```

```{r}
# Smooth the edges of the simulated heights by taking more samples
sim.height <- sim(m4.3, data=list(weight=weight.seq), n=1e4)
height.PI <- apply(sim.height, 2, PI, prob=0.89)
```

```{r}
# `sim` under the hood
post <- extract.samples(m4.3)
weight.seq <- 25:70
sim.height <- sapply(weight.seq, function(weight)
  rnorm(
    n=nrow(post),
    mean=post$a + post$b*(weight-xbar),
    sd=post$sigma
  ))
height.PI <- apply(sim.height, 2, PI, prob=0.89)
```

## 4.5 Curves from lines

### 4.5.1 Polynomial regression

```{r 4.64}
# load in full dataset
library(rethinking)
data("Howell1")
d <- Howell1
plot(height~weight, data=d)
```

Quadratic model:

$$ h_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \beta_{i}x_{i} + \beta_{2}x^2_{i} $$
$$ \alpha \sim \text{Normal}(178, 20) $$
$$ \beta_{1} \sim \text{Log-Normal}(0, 1) $$
$$ \beta_{2} \sim \text{Normal}(0, 1) $$
$$ \sigma \sim \text{Uniform}(0,50) $$

```{r 4.65}
# Standardize weight variables and fit a polynomial model
d$weight_s <- (d$weight - mean(d$weight))/sd(d$weight)
d$weight_s2 <- d$weight_s^2
m4.5 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1*weight_s + b2*weight_s2,
    a ~ dnorm(178, 20),
    b1 ~ dlnorm(0, 1),
    b2 ~ dnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data=d
)

```

```{r 4.66}
# look at table of coefficients
precis(m4.5)

```

```{r 4.67}
# Get data for plotting the posteriors
weight.seq <- seq(from=-2.2, to=2, length.out=30)
pred_dat <- list(weight_s=weight.seq, weight_s2=weight.seq^2)
mu <- link(m4.5, data=pred_dat)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob=0.89)
sim.height <- sim(m4.5, data=pred_dat)
height.PI <- apply(sim.height, 2, PI, prob=0.89)
```

```{r 4.68}
# plot the posterior
plot(height ~ weight_s, d, col=col.alpha(rangi2,0.5))
lines(weight.seq, mu.mean)
shade(mu.PI, weight.seq)
shade(height.PI, weight.seq)
```

Cubic model:

$$ h_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \beta_{i}x_{i} + \beta_{2}x^2_{i} + \beta_{3}x_{i}^3$$
$$ \alpha \sim \text{Normal}(178, 20) $$
$$ \beta_{1} \sim \text{Log-Normal}(0, 1) $$
$$ \beta_{2} \sim \text{Normal}(0, 10) $$
$$ \beta_{3} \sim \text{Normal}(0, 10) $$
$$ \sigma \sim \text{Uniform}(0,50) $$

```{r 4.69}
# Fit Cubic model

d$weight_s3 <- d$weight_s^3

m4.6 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1*weight_s + b2*weight_s2 + b3*weight_s3,
    a ~ dnorm(178, 20),
    b1 ~ dlnorm(0, 1),
    b2 ~ dnorm(0, 10),
    b3 ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data=d
)
```


```{r 4.70-4.71}
# plot data on original scale
plot(height ~ weight_s, d, col=col.alpha(rangi2,0.5), xaxt="n")
at <- c(-2, -1, 0, 1, 2)
labels <- at*sd(d$weight) + mean(d$weight)
axis(side=1, at=at, labels=round(labels,1))
```

### 4.5.2 Splines

```{r 4.72}
# Load in cherry blossom data
library(rethinking)
data("cherry_blossoms")
d <- cherry_blossoms
precis(d)
```

```{r}
plot(doy~year, data=d)
```

```{r 4.73}
# Create knots for B-splines
d2 <- d[complete.cases(d$doy), ] # complete cases on doy
num_knots <- 15
knot_list <- quantile(d2$year, probs=seq(0, 1, length.out=num_knots))
```

```{r 4.74}
# Create basis functions for cubic splines
library(splines)
B <- bs(d2$year,
        knots=knot_list[-c(1, num_knots)],
        degree=3, intercept=TRUE)
```


```{r 4.75}
# Plot basis functions
plot(NULL, xlim=range(d2$year), ylim=c(0,1), xlab="year", ylab="basis")
for (i in 1:ncol(B)) lines(d2$year, B[,i])
```

B-Spline model:

$$ D_{i} \sim \text{Normal}(\mu_{i}, \sigma) $$
$$ \mu_{i} = \alpha + \sum_{k=1}^{K} w_{k}B_{k,i} $$

$$ \alpha \sim \text{Normal}(100, 10) $$
$$ w_{j} \sim \text{Normal}(0, 10) $$
$$ \sigma \sim \text{Exponential}(1) $$

```{r 4.76}
# Define and fit B-spline model
m4.7 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + B %*% w,
    a ~ dnorm(100, 10),
    w ~ dnorm(0, 10),
    sigma ~ dexp(1)
  ),
  data=list( D=d2$doy, B=B),
  start=list(w=rep(0, ncol(B)))
)
```

```{r}
precis(m4.7, depth=2)
```

```{r 4.77}
# Get weighted basis functions
post <- extract.samples(m4.7)
w <- apply(post$w, 2, mean)
plot(NULL, xlim=range(d2$year), ylim=c(-6,6),
     xlab="year", ylab="basis * weight")
for(i in 1:ncol(B)) lines(d2$year, w[i]*B[,i])
```

```{r 4.78}
# Plot 97% posterior interval for mu at each year
mu <- link(m4.7)
mu_PI <- apply(mu,2,PI,0.97)
plot(d2$year, d2$doy, col=col.alpha(rangi2,0.3), pch=16)
shade(mu_PI, d2$year, col=col.alpha("black", 0.5))
```









