---
title: "Chapter_6"
author: "Trevor Williams"
date: "2026-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```

## Introduction

```{r 6.1}
# Overthinking simulated science distortion
set.seed(1914)
N <- 200 # Num grant proposals
p <- 0.1 # proprotion to select
# uncorrelated newsworthiness and trustworthiness
nw <- rnorm(N)
tw <- rnorm(N)
# Select top 10% of combined scores
S <- nw + tw # Total score
q <- quantile(S, 1-p) # top 10% threshold
selected <- ifelse(S >= q, TRUE, FALSE)
cor(tw[selected], nw[selected])
```

## 6.1 Multicollinearity

### 6.1.1 Multicollinear legs

```{r 6.2}
# Simulate height and leg lengths
N <- 100 # number of individuals
set.seed(909)
height <- rnorm(N, 10, 2) # sim total height of each 
leg_prop <- runif(N, 0.4, 0.5)
leg_left <- leg_prop*height + rnorm(N, 0, 0.02) # sim left leg as proportion + error
leg_right <- leg_prop*height + rnorm(N, 0, 0.02) # sim right leg as proportion + error
d <- data.frame(height, leg_left, leg_right) # Combine into data frame
```

```{r 6.3}
# Get posterior approximation of model predicting height by leg lengths
m6.1 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl*leg_left + br*leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.1)
```

```{r 6.4}
# get graphical view of parameter posteriors
plot(precis(m6.1))
```

```{r 6.5}
# Examine joint posterior distribution of br and bl
post <- extract.samples(m6.1)
plot(bl ~ br, post, col=col.alpha(rangi2,0.1), pch=16)
```

```{r 6.6}
# Get posterior distribution of sum of bl and br
sum_blbr <- post$bl + post$br
dens(sum_blbr, col=rangi2, lwd=2, xlab="sum of bl and br")
```

```{r 6.7}
# Approximate model with only one leg length and compare to summed posterior above
m6.2 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl*leg_left,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.2) 
```

### 6.1.2 Multicollinear milk

```{r 6.8}
# Read in and prep milk data
library(rethinking)
data(milk)
d <- milk
d$K <- standardize(d$kcal.per.g)
d$F <- standardize(d$perc.fat)
d$L <- standardize(d$perc.lactose)
```


```{r 6.9}
# Do two bivariate regressions for perc.fat and perc.lactose

# kcal.per.g regressed on perc.fat
m6.3 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bF*F,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

# kcal.per.g regressed on perc.lactose
m6.4 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bL * L,
    a ~ dnorm(0, 0.2),
    bL ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d
)

precis(m6.3)
precis(m6.4)
```

```{r 6.10}
# Model with both predictors
m6.5 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bF*F + bL*L,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    bL ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d
)

precis(m6.5)

```

```{r 6.11}
# Visualize correlation using pairs
pairs(~ kcal.per.g + perc.fat + perc.lactose, data=d, col=rangi2)
```

```{r 6.12}
# Overthinking: Simulating collinearity
data(milk)
d <- milk
sim.coll <- function(r=0.9) {
  d$x <- rnorm(nrow(d), mean=r*d$perc.fat, 
  sd=sqrt((1-r^2)*var(d$perc.fat)))
  m <- lm(kcal.per.g ~ perc.fat + x, data=d)
  sqrt(diag(vcov(m)))[2] # stddev of parameter
}
rep.sim.coll <- function(r=0.9, n=100) {
  stddev <- replicate(n, sim.coll(r))
  mean(stddev)
}
r.seq <- seq(from=0, to=0.99, by=0.01)
stddev <- sapply(r.seq, function(z) rep.sim.coll(r=z, n=100))
plot(stddev~r.seq, type="l", col=rangi2, lwd=2, xlab="correlation")
```

## 6.2 Post-treatment bias

```{r 6.13}
# Simulate and examine post-treatment effect

set.seed(71)
# number of plants
N <- 100

# simulate initial heights
h0 <- rnorm(N,10,2)

# assign treatments and simulate fungus and growth
treatment <- rep(0:1, each=N/2)
fungus <- rbinom(N, size=1, prob=0.5-treatment*0.5)
h1 <- h0 + rnorm(N, 5-3*fungus)

# compose a clean data frame
d <- data.frame(h0=h0, h1=h1, treatment=treatment, fungus=fungus)
precis(d)
```
### 6.2.1 A prior is born

Model before taking predictors into account:

$$ h_{1,i} \sim \text{Normal}(\mu, \sigma) $$
$$ \mu_{i} = h_{0,i}*p $$

```{r 6.14}
# Create and expect prior samples for the proportion of h1/h0
sim_p <- rlnorm(1e4, 0, 0.25)
precis(data.frame(sim_p))
```

```{r 6.15} 
# Fit model without predictors
m6.6 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- h0*p,
    p ~ dlnorm(0, 0.25),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.6)
```


Model with predictors:

$$ h_{1,i} \sim \text{Normal}(\mu, \sigma) $$
$$ \mu_{i} = h_{0,i}*p $$
$$ p = \alpha + \beta_{T}T_{i} + \beta_{F}F_{i} $$
$$ \alpha \sim \text{Log-Normal}(0,0.25) $$
$$ \beta_{T} \sim \text{Normal}(0, 0.5) $$
$$ \beta_{F} \sim \text{Normal}(0, 0.5) $$
$$ \sigma \sim \text{Exponential}(1) $$

```{r 6.16}
# Fit the model with both treatment and fungus
m6.7 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- h0 * p,
    p <- a + bt*treatment + bf*fungus,
    a ~ dlnorm(0, 0.2),
    bt ~ dnorm(0, 0.5),
    bf ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.7)
```

### 6.2.2 Blocked by consequence


```{r 17}
# Fit model without fungus - answers question what is the effect of treatment on growth
m6.8 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- h0 * p,
    p <- a + bt*treatment,
    a ~ dlnorm(0, 0.2),
    bt ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.8)
```

### 6.2.3 Fungus and d-separation

```{r 6.18}
# Look at dag for problem above
library(dagitty)
plant_dag <- dagitty("dag {
                     H_0 -> H_1
                     F -> H_1
                     T -> F
                     }")
coordinates(plant_dag) <- list(x=c(H_0=0,T=2,F=1.5,H_1=1),
                               y=c(H_0=0,T=0,F=0,H_1=0))
drawdag(plant_dag)
```

```{r 6.19}
# Get implied conditional independencies for DAG
impliedConditionalIndependencies(plant_dag)
```

```{r 6.20}
# Uncouple T from H1 and add M affecting both H1 and F
set.seed(71)
N <- 1000
h0 <- rnorm(N,10,2)
treatment <- rep(0:1, each=N/2)
M <- rbern(N)
fungus <- rbinom(N, size=1, prob=0.5 - treatment*0.4 + 0.4*M)
h1 <- h0 + rnorm(N, 5 + 3*M)
d2 <- data.frame(h0=h0, h1=h1, treatment=treatment, fungus=fungus)

m6.7b <- quap(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- h0 * p,
    p <- a + bt*treatment + bf*fungus,
    a ~ dlnorm(0, 0.2),
    bt ~ dnorm(0, 0.5),
    bf ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d2
)


m6.8b <- quap(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- h0 * p,
    p <- a + bt*treatment,
    a ~ dlnorm(0, 0.2),
    bt ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data=d2
)

precis(m6.7)
precis(m6.8)

```

## 6.3 Collider bias

Example of collider Science and trustworthiness/newsworthiness

```{r}
science_dag <- dagitty(" dag { T -> S <- N } ")
coordinates(science_dag) = list(x=c(T=0, S=1, N=2),
                                y=c(T=0, S=0, N=0))
drawdag(science_dag)
```

### 6.3.1 Collider of false sorrow

DAG:

```{r}
happiness_dag <- dagitty(" dag { H -> M <- A } ")
coordinates(happiness_dag) = list(x=c(H=0, M=1, A=2),
                                y=c(H=0, M=0, A=0))
drawdag(happiness_dag)
```

```{r 6.21}
# Simulate age happiness agent-based model
d <- sim_happiness(seed=1977, N_years=1000)
precis(d)
```

```{r 6.22}
# filter to adults and rescale age
d2 <- d[d$age > 17, ] # only adults
d2$A <- (d2$age - 18)/(65-18)
```

```{r 6.23}
# Fit the posterior of model containing marriage as a predictor
d2$mid <- d2$married + 1
m6.9 <- quap(
  alist(
    happiness ~ dnorm(mu, sigma),
    mu <- a[mid] + bA*A,
    a[mid] ~ dnorm(0, 1),
    bA ~ dnorm(0,2),
    sigma ~ dexp(1)
  ),
  data=d2
)
precis(m6.9, depth=2)
```
```{r 6.24}
# approximate posterior from model without marriage as a predictor
m6.10 <- quap(
  alist(
    happiness ~ dnorm(mu, sigma),
    mu <- a + bA*A,
    a ~ dnorm(0, 1),
    bA ~ dnorm(0, 2),
    sigma ~ dexp(1)
  ),
  data=d2
)
precis(m6.10)
```

### 6.3.2 The huanted DAG

Education DAG without unmeasured variable

```{r}
education_dag1 <- dagitty(" dag { 
                          G -> P -> C
                          G -> C } ")
coordinates(education_dag1) <- list(x=c(G=1, P=2, C=2),
                                   y=c(G=-1, P=-1, C=0))
drawdag(education_dag1)
```

DAG with unobserved huanted U

```{r}
education_dag1 <- dagitty(" dag { 
                          G -> P -> C
                          G -> C 
                          [Latent] U -> P
                          [Latent] U -> C} ")
coordinates(education_dag1) <- list(x=c(G=1, P=2, C=2, U=3),
                                   y=c(G=-1, P=-1, U=-0.5, C=0))
drawdag(education_dag1)
```

```{r 6.25}
# Set associations and parameters for education simulation
N <- 200 # Number of grandparent-parent-child triads
b_GP <- 1 # Direct effect of G on P
b_GC <- 0 # Direct effect of G on C
b_PC <- 1 # Direct effect of P on C
b_U <- 2 # Direct effect of U on P and C
```

```{r 6.26}
# Use slopes to to draw random observations
set.seed(1)
U <- 2*rbern(N, 0.5) -1
G <- rnorm(N)
P <- rnorm(N, b_GP*G + b_U*U)
C <- rnorm(N, b_PC*P + b_GC*G + b_U*U)
d <- data.frame(C=C, P=P, G=G, U=U)
d

```

```{r 6.27}
# fit model with G + P affect on C
m6.11 <- quap(
  alist(
    C ~ dnorm(mu, sigma),
    mu <- a + b_PC*P + b_GC*G,
    a ~ dnorm(0,1),
    c(b_PC,b_GC) ~ dnorm(0,1),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.11)
```

```{r 6.28}
# Now fit with U
m6.12 <- quap(
  alist(
    C ~ dnorm(mu, sigma),
    mu <- a + b_PC*P + b_GC*G + b_U*U,
    a ~ dnorm(0, 1),
    c(b_PC, b_GC, b_U) ~ dnorm(0,1),
    sigma ~ dexp(1)
  ),
  data=d
)
precis(m6.12)
```

## 6.4 Confronting confounding

### 6.4.1 Shutting the backdoor

4 Types of Relations in DAGS

1. The Fork

```{r}
fork <- dagitty("dag { Z -> X
                Z -> Y} ")
coordinates(fork) <- list(x=c(X=0, Z=1, Y=2),
                          y=c(X=-1, Z=0, Y=-1))
drawdag(fork)
```
  - Conditioning on Z closes the path
  
2. Pipe

```{r}
fork <- dagitty("dag { X -> Z
                Z -> Y} ")
coordinates(fork) <- list(x=c(X=0, Z=1, Y=2),
                          y=c(X=-1, Z=0, Y=1))
drawdag(fork)
```

  - conditioning on Z closes the path
  
3. Collider
```{r}
fork <- dagitty("dag { X -> Z
                Y -> Z} ")
coordinates(fork) <- list(x=c(X=0, Z=1, Y=2),
                          y=c(X=0, Z=-1, Y=0))
drawdag(fork)
```

  - Conditioning on Z opens the path
  
4. Descendent

```{r}
fork <- dagitty("dag { X -> Z
                Y -> Z
                Z -> D } ")
coordinates(fork) <- list(x=c(X=0, Z=1, Y=2, D=1),
                          y=c(X=0, Z=-1, Y=0, D=0))
drawdag(fork)
```

Conditioning on D, conditions (to a lesser extend) on Z - Opens path from X to Y

### 6.4.2

```{r 6.29}
dag_6.1 <- dagitty("dag { 
  U [unobserved]
  X -> Y
  X <- U <- A -> C -> Y
  U -> B <- C
                } ")
#coordinates(fork) <- list(x=c(X=0, U=0, A=1, B=1, C=2, Y=2),
#                          y=c(X=0, U=-1, Y=0, B=-0.5, C=-1, A=-1.5))
drawdag(fork)
adjustmentSets(dag_6.1, exposure="X", outcome="Y")
```

### 6.4.3 Backdoor waffles

```{r 6.30}
dag_6.2 <- dagitty("dag {
  A -> D
  A -> M -> D
  A <- S -> M
  S -> W -> D
                   }")
drawdag(dag_6.2)
adjustmentSets(dag_6.2, exposure="W", outcome="D")
```

```{r 6.31}
impliedConditionalIndependencies(dag_6.2)
```

