---
title: "Chapter_3_Questions"
author: "Trevor Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```

## Easy Questions

The Easy problems use the samples from the posterior distribution for the globe tossing example. This code will give you a specific set of samples so that you can check your answers exactly

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size=9, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
```

**3E1. How much posterior probability liees below p = 0.2?**

```{r}
sum(samples < 0.2) / 10000
```

0.04% of the posterior probability lies below 0.2.

**3E2. How much posterior probability lies above p = 0.8?**

```{r}
sum(samples > 0.8) / 10000
```

11.16% of the posterior probability lies above 0.8

**3E3. How much posterior probability lies between p=0.2 and p=0.8?**

```{r}
sum(samples > 0.2 & samples < 0.8) / length(samples)
```

88.8% of the posterior probability lies between 0.2 and 0.8

**3E4. 20% of the posterior probability lies below which value of p?**

```{r}
quantile(samples, 0.2)
```

20% of the posterior probability lies below 0.519

**3E5. 20% of the posterior probability lies above which value of p?**

```{r}
quantile(samples, 0.8)
```

20% of the posterior probability lies above 0.756

**3E6. Which values of P contain the narrowest interval equal to 66% of the posterior probability?**

```{r}
HPDI(samples, 0.66)
```

The narrowest interval equal to 66% of the posterior probability is between 0.509 and 0.774

**3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?**

```{r}
PI(samples, 0.66)
```

The middle 66% of p lies between 0.503 and 0.770

## Medium Questions

**3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before**

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1, length(p_grid))
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot(p_grid, posterior)
```

**3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for $p$**

```{r}
samples <- sample(p_grid, prob=posterior, size=10000, replace=TRUE)
HPDI(samples, prob=0.9)
```

**3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in $p$. What is the probability of observing 8 water in 15 tosses?**

```{r}
# Create posterior predictive distribution

w <- rbinom(1e4, size=15, prob=samples)
simplehist(w)


# Get the probability of observing 8 water in 15 tosses
sum(w == 8) / length(w)
```

The probabililty of observing 8 water in 15 tosses is approximately 14.87%

**3M4. using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses**

```{r}
likelihood_6of9 <- dbinom(6, size=9, prob=p_grid)
prior_6of9 <- posterior
(posterior_6of9 <- sum(likelihood_6of9 * prior_6of9))

# with sampling
set.seed(100)
w_6of9 <- rbinom(10000, size=9, prob=samples)
simplehist(w_6of9)
mean(w_6of9 == 6)
```

**3M5. Start over at 3M1, but now use a prior that is zero below p=0.5 and a constant above p=0.5. This corresponds to prior information that a majority of the earth's surface is water. Repeat each problem above and compare the inferences. What difference does teh better prior make? If it helps, compare inferences (using both priors) to the true value p=0.7**

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior_new <- ifelse(p_grid < 0.5, 0, 1)
likelihood_new <- dbinom(8, size=15, prob=p_grid)
posterior_new <- likelihood_new * prior_new
posterior_new <- posterior_new / sum(posterior_new)

set.seed(100)
samples_new <- sample(p_grid, prob=posterior_new, size=1e4, replace=TRUE)
plot(p_grid, posterior_new, type='l')
```

90% HPDI
```{r}
HPDI(samples_new, prob=0.9)
```

The 90% HPDI is shifted on the lower end to start at 0.5, and is narrower.

posterior predictive check
```{r}
w_new <- rbinom(1e4, size=15, prob=samples_new)
simplehist(w_new)

sum(w_new == 8) / length(w_new)
```
The probability of seeing 8 waters in 15 tosses increased slightly


```{r}

w_new <- rbinom(1e4, size=9, prob=samples_new)
simplehist(w_new)

sum(w_new == 6) / length(w_new)
```

The probability of seeing 6 out of 9 tosses also increased a little.

```{r}
par(mfrow=c(1, 2))
hist(samples)
hist(samples_new)
par(mfrow=c(1,1))
```

**3M6. Suppose you want to estimate the Earth's proportion of water very precisely. Specifically, you want the 99th percentile interval of the posterior distribution of $p$ to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this?**

```{r}
f <- function(N) {
  p_true <- 0.7
  W <- rbinom(1, size=N, prob=p_true)
  p_grid <- seq(from=0, to=1, length.out=1000)
  prior <- rep(1, length(p_grid))
  likelihood <- dbinom(W, size=N, prob=p_grid)
  posterior <- likelihood * prior
  posterior <- posterior / sum(posterior)
  samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
  PI99 <- PI(samples, .99)
  as.numeric(PI99[2]-PI99[1])
}
Nlist <- c(20, 50, 100, 200, 500, 1000, 2000, 3000)
Nlist <- rep(Nlist, each = 100)
width <- sapply(Nlist, f)
plot(Nlist, width)
abline(h = 0.05, col='red')

```


