---
title: "Statistical Rethinking Chapter 3: Sampling the Imaginary"
output: html_notebook
---

```{r setup, include=FALSE}

library(rethinking)

knitr::opts_knit$set(root.dir = "~/dev/SRBayesianR/")


```

### Introduction

Vampire example:

$$ Pr(\text{vampire|positive}) = \frac{Pr(\text{positive|vampire})Pr(\text{vampire})}{Pr(\text{positive|vampire})+Pr(\text{positive|mortal})(1-Pr(\text{vampire}))}$$
$$ =\frac{0.95*0.001}{(0.95*0.001) + (0.01*(1-0.001))}$$
Code to calculate:

```{r}
Pr_positive_vampire <- 0.95
Pr_positive_mortal <- 0.01
Pr_vampire <- 0.001
Pr_Positive <- Pr_positive_vampire*Pr_vampire + Pr_positive_mortal*(1-Pr_vampire)
( Pr_vampire_positive <- Pr_positive_vampire*Pr_vampire/Pr_Positive)
```

Shows that when the condition of interest is very rare, having a test that finds the true cases is no guarantee that the positive result carries much information since most are false positives

Can use counts to display the same information:

 1. In a population of 10000 people, 100 are vampires
 2. of the 100 who are vampires, 95 will test positive for vampirism
 3. Of the 99,900 mortals, 999 will test positive for vampirism
 
$$ Pr(\text{vampire|positive})=\frac{95}{1094} = 0.087 $$
This is called the *frequency format* or *natural frequencies*. 

Thinking Bayesian -> we can exploit this format by taking probability distributions and sampling from them to produce counts

* Can also draw samples from the posterior distribution -> sampled events are parameter values
 
### 3.1 Sampling from a grid-approximate posterior

Before we sample from the posterior, we need to make one. Using the Globe tossing model and grid approximation let's create the posterior:

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prob_p <- rep(1, 1000) # prior
prob_data <- dbinom(6, size=9, prob=p_grid)
posterior <- prob_data * prob_p
posterior <- posterior/sum(posterior)

```


Now we want to draw 10000 samples from this posterior
* Imagine the posterior is a bucket full of parameters values, with each value existing in proportion to the posterior probability in `posterior`
* We can sample from this bucket to approximate the posterior

```{r}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)

```

Lets plot the samples and the *density estimate* using the `dens()` function from `rethinking`

```{r}
par(mfrow=c(1,2))
plot(samples, col=rgb(0, 0, 1, alpha = 0.5), pch=16, ylim=c(0,1))
dens(samples, col='blue')
```

### 3.2 Sampling to summarize

How do you summarize the posterior? Common questions include:

1. Questions about intervals of *defined boundaries*
  * How much posterior probability lies below some parameter value?
  * How much posterior probability lies between two parameters values?

2. Questions about intervals of *defined probability mass*
  * Which parameter value marks the lower 5% of the posterior probability?
  * Which range of parameter values contains 90% of the posterior probability?

3. Questions about *point estimates*
  * Which parameter value has highest posterior probability?

#### 3.2.1 Intervals of defined boundaries

What is the posterior probability that the proportion of water is less than 0.5

```{r}
# Add up posterior probability where p < 0.5
sum(posterior[p_grid<0.5])
```

This is the way to do it for single parameters with grid approximation -> trickier when more parameters present

The same calculation using samples from the posterior can be computed as the proportion (or frequency) of samples that match your criteria:

```{r}
sum(samples < 0.5)/length(samples)
```

This is generalizable to instances with more than one parameter as well. It can also be used to find how much of the posterior probability lies between 0.5 and 0.75:

```{r}
sum(samples > 0.5 & samples < 0.75)/length(samples)
```

#### 3.2.2 Intervals of defined mass

Intervals of defined mass - or *confidence intervals* or *credibility intervals* or *compatibility intervals* - are ranges of parameter values compatible with the model and data.

 * Report two parameter values that contain between them a specified amount of posterior probability, which is a probability mass.
 
To find the lower 80% posterior probability use `quantile()`

```{r}
quantile(samples, 0.8)
```

To find the middle 80% find the parameter values at the 10th and 90th percentiles

```{r}
quantile( samples, c(0.1, 0.9))
```

These types of intervals, where equal probability mass is assigned to each tail, can also be called *percentile intervals (PI)*. However, it isn't perfect, consider the following posterior distribution:

```{r}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1,1000)
likelihood <- dbinom(3, size=3, prob=p_grid)
posterior <- likelihood*prior
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, size=1e4, replace=TRUE, prob=posterior)

par(mfrow=c(1,2))

plot(samples, col=rgb(0, 0, 1, alpha = 0.5), pch=16, ylim=c(0,1))
dens(samples, col='blue')
```


You can use `PI()` from `rethinking` to calculate middle intervals:

```{r}
PI(samples, prob=0.5)
```

This puts 25% of the probability mass above and below the interval, but excludes the most probable parameter values near p=1, misleading information on the shape of the distribution.

Instead you can use the *highest posterior density interval* (HPDI) - the narrowest interval containing the specified probability mass. You can calculate it using `HPDI()` from `rethinking`

```{r}
HPDI(samples, prob=0.5)
```

Disadvantages of HPDI:
 1. More computationally intensive
 2. Suffers from greater simulation variance
 3. Harder to understand